Reaction-diffusion spatial modeling of COVID-19 in Chicago

### 2021-11-12 update-10-22/update-10-22.tex
Non-dimensional parameters and equations

	We begin by reducing the model.
	First, we ignore non-COVID deaths, so $\mu = 0$.
	Next we introduce the fractions
	\begin{align*}
		s &= \frac{S}{N}, & e &= \frac{E}{N}, & ar &= \frac{AR}{N}, & a &= \frac{A}{N}, \\
		i &= \frac{I}{N}, & h &= \frac{H}{N}, & r &= \frac{R}{N}, & d &= \frac{D}{N}.
	\end{align*}
	With these fractions and $\mu = 0$, we have the conservation law
	\begin{equation} \label{eq:conservation}
		s + e + ar + a + i + h + r + d = 1.
	\end{equation}
	The ODE system becomes
	\begin{align*}
		s_t &=	 - \beta_{A} a s - \beta_{I} i s, \\
		e_t	&=	\beta_{A} a s + \beta_{I} i s - (\sigma_A + \sigma_I) e, \\
		ar_t &= M_{AR} a, \\
		a_t	&=	\sigma_A e - M_{AR} a, \\
		i_t	&=	\sigma_I e - M i, \\
		h_t	&=	\gamma M i - (1 - \omega) \chi h - \omega \psi h, \\
		r_t	&=	(1 - \gamma) M i + (1 - \omega) \chi h, \\
		d_t	&=	\omega \psi h.
	\end{align*}
	At this point it is worth noting the dimensions of the components of the system.
	Since $[X] = \mathrm{population}$ where $X \in \{N, S, E, AR, A, I, H, R, D\}$, then $[x] = 1$ where $x \in \{s, e, ar, a, i, h, r, d\}$.
	Then,
	$$[\beta_A] = [\beta_I] = [\sigma_A] = [\sigma_I] = [M_{AR}] = [M] = [\chi] = [\psi] = \frac{1}{T}$$
	and $[\omega] = [\gamma] = 1$.
	We introduce the non-dimensional time variable $\tau = t M$.
	Then by the chain rule, $x_t = M x_\tau$.
	Next, we define the following non-dimensional parameters
	\begin{align*}
		\alpha_A &= \frac{\beta_A}{M}, & \alpha_I &= \frac{\beta_I}{M}, & \lambda_A &= \frac{\sigma_A}{M}, & \lambda_I &= \frac{\sigma_I}{M}
	\end{align*}

	Using these non-dimensional parameters and the conservation law in Equation \ref{eq:conservation}, we can replace the ODE system with
	\begin{align}
		s_\tau &= - \alpha_A a s - \alpha_I i s, \\
		e_\tau &= \alpha_A a s + \alpha_I i s - (\lambda_A + \lambda_I) e, \\
		ar_\tau &= \xi a, \\
		a_\tau &= \lambda_A e - \xi a, \\
		i_\tau &= \lambda_I e - i
	\end{align}

@@

	To account for changes in virus transmission due to the shelter-in-place order, we impose a time dependence on the transmission rates $\beta$ as in Equations (\ref{eq:beta-si}) and (\ref{eq:beta-sa}).
	\begin{align}
		\label{eq:beta-si}
		\beta_{SI} (t) &= \beta_{SI} \left( \eta_{SI} + (1 - \eta_{SI}) \frac{1 - \tanh [2(t - t_q)]}{2} \right) \\
		\label{eq:beta-sa}
		\beta_{SA} (t) &= \beta_{SA} \left( \eta_{SA} + (1 - \eta_{SA}) \frac{1 - \tanh [2(t - t_q)]}{2} \right)
	\end{align}

	\begin{table}[h]
		\centering
		\caption{ODE parameters: optimal (best-fitting), median and interquartile range, and variation range used in the optimization algorithm.
			Initial parameter guesses were uniformly sampled within these ranges.}
		\label{tab:parameters}
		\begin{tabular}{ c c c c }
			\hline
			\hline
																								&					&	Median (IQR)	&	Initial value \\
			\hline
			Transmission rate, $S \to I$							&	$\beta_{SI}$	&									&	$c \in \U [0,1]$ \\
			Transmission rate, $S \to A$							&	$\beta_{SA}$	&									& 	$c \in \U [0,1]$ \\
			Lockdown effect, $S \to I$															&	$\eta_{SI}$ 	&									& 	$c \in \U [0,1]$ \\
			Lockdown effect, $S \to A$															&	$\eta_{SA}$		&									& 	$c \in \U [0,1]$ \\
			Incubation period, $E \to I$ [days]													&	$1 / \sigma_I$	&									& 	$1 / k, k \in \U [2,7]$ \\
			Latent period, $E \to A$ [days]														&	$1 / \sigma_A$	&									& 	$1 / k, k \in \U [2,7]$ \\
			Infectivity period [days]															&	$1 / M$			&									& 	$1 / k, k \in \U [5,12]$ \\
			Recovery period, $A \to AR$ [days]													&	$1 / M_{AR}$	&									& 	$1 / k, k \in \U [5,12]$ \\
			Recovery period, $H \to R$ [days]													&	$1 / \chi$		&									& 	$1 / k, k \in \U [5,20]$ \\
			Period to deceased, $H \to D$ [days]												&	$1 / \psi$		&									& 	$1 / k, k \in \U [5,20]$ \\
			Conversion fraction $(I \xrightarrow{\gamma} H, \ I \xrightarrow{1 - \gamma} R)$	&	$\gamma$		&									& 	$c \in \U [0.25,0.75]$ \\
			Conversion fraction $(H \xrightarrow{\omega} D, \ H \xrightarrow{1 - \omega} R)$	&	$\omega$		&									& 	$c \in \U [0.1,0.5]$ \\
			Initial population fraction, exposed												&	$E_0 / I_0$		&									& 	$c \in \U [1,5]$ \\
			Initial population fraction, asymptomatic											&	$A_0 / I_0$		&									& 	$c \in \U [1,5]$ \\
			\hline
			\hline
		\end{tabular}
	\end{table}

>>>>

	We begin by reducing the model.
	First, we ignore non-COVID deaths, so $\mu = 0$.
	Next we introduce the fractions
	\begin{align*}
		s &= \frac{S}{N}, & e &= \frac{E}{N}, & ar &= \frac{AR}{N}, & a &= \frac{A}{N}, \\
		i &= \frac{I}{N}, & h &= \frac{H}{N}, & r &= \frac{R}{N}, & d &= \frac{D}{N}.
	\end{align*}
	With these fractions and $\mu = 0$, we have the conservation law
	\begin{equation} \label{eq:conservation}
		s + e + ar + a + i + h + r + d = 1.
	\end{equation}
	The ODE system becomes
	\begin{align*}
		s_t &=	 - \beta_{A} a s - \beta_{I} i s, \\
		e_t	&=	\beta_{A} a s + \beta_{I} i s - (\sigma_A + \sigma_I) e, \\
		ar_t &= M_{AR} a, \\
		a_t	&=	\sigma_A e - M_{AR} a, \\
		i_t	&=	\sigma_I e - M i, \\
		h_t	&=	\gamma M i - (1 - \omega) \chi h - \omega \psi h, \\
		r_t	&=	(1 - \gamma) M i + (1 - \omega) \chi h, \\
		d_t	&=	\omega \psi h.
	\end{align*}
	At this point it is worth noting the dimensions of the components of the system.
	Since $[X] = \mathrm{population}$ where $X \in \{N, S, E, AR, A, I, H, R, D\}$, then $[x] = 1$ where $x \in \{s, e, ar, a, i, h, r, d\}$.
	Then,
	$$[\beta_A] = [\beta_I] = [\sigma_A] = [\sigma_I] = [M_{AR}] = [M] = [\chi] = [\psi] = \frac{1}{T}$$
	and $[\omega] = [\gamma] = 1$.
	We introduce the non-dimensional time variable $\tau = t M$.
	Then by the chain rule, $x_t = M x_\tau$.
	Next, we define the following non-dimensional parameters
	\begin{align*}
		\alpha_A &= \frac{\beta_A}{M}, & \alpha_I &= \frac{\beta_I}{M}, & \lambda_A &= \frac{\sigma_A}{M}, & \lambda_I &= \frac{\sigma_I}{M}, \\
		\eta &= \frac{M_{AR}}{M}, & \theta &= \frac{\chi}{M}, & \kappa &= \frac{\psi}{M}.
	\end{align*}

	Using these non-dimensional parameters and the conservation law in Equation \ref{eq:conservation}, we can replace the ODE system with
	\begin{align}
		\label{eq:s_t}
		s_\tau &= - \alpha_A a s - \alpha_I i s, \\
		\label{eq:e_t}
		e_\tau &= \alpha_A a s + \alpha_I i s - (\lambda_A + \lambda_I) e, \\
		\label{eq:ar_t}
		ar_\tau &= \xi a, \\
		\label{eq:a_t}
		a_\tau &= \lambda_A e - \xi a, \\
		\label{eq:i_t}
		i_\tau &= \lambda_I e - i, \\
		\label{eq:r_t}
		r_\tau &= (1 - \gamma) i + (1 - \omega) \theta h.
	\end{align}
	Then by the conservation law, we have
	\begin{equation} \label{eq:d_t}
		d = n - (s + e + ar + a + i + r).
	\end{equation}

@@

	To account for changes in virus transmission due to the shelter-in-place order, we impose a time dependence on the transmission rates $\beta$ as in Equations (\ref{eq:beta-si}) and (\ref{eq:beta-sa}).
	\begin{align}
		\label{eq:beta-si}
		\beta_{I} (t) &= \beta_{I} \left( \eta_{I} + (1 - \eta_{I}) \frac{1 - \tanh [2(t - t_q)]}{2} \right) \\
		\label{eq:beta-sa}
		\beta_{A} (t) &= \beta_{A} \left( \eta_{A} + (1 - \eta_{A}) \frac{1 - \tanh [2(t - t_q)]}{2} \right)
	\end{align}

	\begin{table}[h]
		\centering
		\caption{ODE parameters: optimal (best-fitting), median and interquartile range, and variation range used in the optimization algorithm.
			Initial parameter guesses were uniformly sampled within these ranges.}
		\label{tab:parameters}
		\begin{tabular}{ c c c c }
			\hline
			\hline
																								&					&	Median (IQR)	&	Initial value \\
			\hline
			Transmission rate, $S \to I$														&	$\beta_{I}$	&									&	$c \in \U [0,1]$ \\
			Transmission rate, $S \to A$														&	$\beta_{A}$	&									& 	$c \in \U [0,1]$ \\
			Lockdown effect, $S \to I$															&	$\eta_{I}$ 	&									& 	$c \in \U [0,1]$ \\
			Lockdown effect, $S \to A$															&	$\eta_{A}$		&									& 	$c \in \U [0,1]$ \\
			Incubation period, $E \to I$ [days]													&	$1 / \sigma_I$	&									& 	$1 / k, k \in \U [2,7]$ \\
			Latent period, $E \to A$ [days]														&	$1 / \sigma_A$	&									& 	$1 / k, k \in \U [2,7]$ \\
			Infectivity period [days]															&	$1 / M$			&									& 	$1 / k, k \in \U [5,12]$ \\
			Recovery period, $A \to AR$ [days]													&	$1 / M_{AR}$	&									& 	$1 / k, k \in \U [5,12]$ \\
			Recovery period, $H \to R$ [days]													&	$1 / \chi$		&									& 	$1 / k, k \in \U [5,20]$ \\
			Period to deceased, $H \to D$ [days]												&	$1 / \psi$		&									& 	$1 / k, k \in \U [5,20]$ \\
			Conversion fraction $(I \xrightarrow{\gamma} H, \ I \xrightarrow{1 - \gamma} R)$	&	$\gamma$		&									& 	$c \in \U [0.25,0.75]$ \\
			Conversion fraction $(H \xrightarrow{\omega} D, \ H \xrightarrow{1 - \omega} R)$	&	$\omega$		&									& 	$c \in \U [0.1,0.5]$ \\
			Initial population fraction, exposed												&	$E_0 / I_0$		&									& 	$c \in \U [1,5]$ \\
			Initial population fraction, asymptomatic											&	$A_0 / I_0$		&									& 	$c \in \U [1,5]$ \\
			\hline
			\hline
		\end{tabular}
	\end{table}
